---
title: "How to estimate local recombination rates with the Marey map method"
output: rmarkdown::html_vignette
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{RecombinationMap}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(ggplot2)
```

The package is loaded with the following command.

```{r setup}
library(EasyMareyMap)
```


# Create a MareyMap object

You can import the example dataset based on linkage maps for the five chromosomes of the plant *Arabidopsis thaliana* [@serin_construction_2017].


The datasets for the 'EasyMareyMap' package are formatted according to the specifications of input files of the MareyMap package [@rezvoy_mareymap:_2007]. It is a tab-separated file with five columns named 'set', 'map, 'mkr', 'phys', 'gen', 'vld'. Genomic coordinates are in basepairs (bp) and genetic distances in cM.


```{r importMap}
m = read.table("../inst/extdata/Arabidopsis_thaliana.txt",
               header = T,
               sep = "\t")
# The marey map is imported into a 'mareyMap' object
df = mareyMap(m)
# names of slots in the 'mareyMap' object
names(df)
# Column names of the input Marey map
names(df$mareyMap)
```



# Quality control

Here is the command to do outlier selection in a visual mode. You just need to draw a region around markers to remove. Left-click to place points of the polygon, then rigth-click to close the region and return to the script. All markers selected are automatically marked as removed (`vld =  FALSE`).

```
df = outlierSelection(df, "1")
```

<!-- TODO Add a photo of mkr selection -->


Alternatively, you can remove a whole chromosome of your dataset with the following command.


```{r removeChromosome}
df = removeChromosome(df, "5")
```


There is a method to `plot` the Marey maps of a `mareyMap` object. Markers kept for further interpolation are in blue. Markers discarded are in red. Here the whole chromosome 5 is discarded.

```{r plot, message=FALSE, warning=FALSE, echo=FALSE, fig.height = 4, fig.width = 6, fig.align="center", fig.cap = "Marey maps of five chromosomes in *Arabidopsis thaliana*. Invalid markers are shown in red."}
plot_marey(df)
```

There is also a method to display a `summary` of a `mareyMap` object.

```{r summary}
summary(df)
```

# Estimate a recombination map

All the procedure to estimate the recombination map is done with the `recombinationMap()` function which returns a `mareyMap` object with an estimated recombination map. A description of the method is available in Brazier and Glémin [@brazierDiversityDeterminantsRecombination2022].

```{r estimateMap}
res = recombinationMap(df, chromosome = "1", method = "loess")
res = recombinationMap(res, chromosome = "2", method = "loess")
res = recombinationMap(res, chromosome = "3", method = "spline")
```


```{r plotInterpolation, message=FALSE, warning=FALSE, echo=FALSE, fig.height = 4, fig.width = 6, fig.align="center", fig.cap = "Marey maps of five chromosomes in *Arabidopsis thaliana* with the Marey interpolation line and a 95% C.I. (invalid markers are shown in red)."}
plot_marey(res)
```



```{r headMap, echo=TRUE, message=FALSE}
head(res$recMap)
```


```{r summaryResults, echo=TRUE, message=FALSE}
summary(res)
```

Finally, the recombination landscape can be plotted with the `recombinationPlot()` function.

```{r plotMap, message=FALSE, warning=FALSE, echo=FALSE, fig.height = 4, fig.width = 8, fig.align="center", fig.cap = "Recombination landscape of *Arabidopsis thaliana*. The gray area is the 95% confidence interval around the mean estimate."}
plot_recombination(res)
```


## Custom or pointwise physical positions

By default, a grid of windows of constant size is mapped along the chromosome. But you can also provide a data frame of custom start/end positions to estimate the recombination rate within.


```{r customWindows}
pos = data.frame(start = seq(20000, 40000000, by = 40000),
                 end = seq(20000, 40000000, by = 40000) + 20000)
res = recombinationMap(res, chromosome = "4", method = "loess", windows = pos)
```

This custom positions can represent any genomic feature (e.g. gene, exon) or marker (e.g. SNP) for which you want to estimate the average recombination rate. However keep in mind that interpolation is at a large scale (i.e. does not detect fine scale variations of recombination rates) and close markers can have very similar values due to the smoothing.



```{r customWindows2}
genes = data.frame(chromosome = "4",
                   start = c(60000, 99000, 123000, 267400, 500000),
                   end = c(65000, 99500, 1234000, 267401, 500001))
pos = data.frame(start = genes$start,
                 end = genes$end)
res = recombinationMap(res, chromosome = "4", method = "loess", windows = pos)

res$windows$`4`
res$recMap[which(res$recMap$map == "4"),]
```


# Interpolate new genetic positions

You may want to change your physical map and interpolate the genetic distances for your new set of markers. It is possible to interpolate these new positions with the `predict` function corresponding to the interpolation method (e.g. `predict.loess` for the `fit.loess` method). I implemented a generic `predictGeneticMap` function to do the job.


It is recommended to re-use the smoothing parameter previously estimated with the automatic calibration procedure.


```{r predictGeneticMap}
# Example for chromosome 1
phys = res$mareyMap$phys[which(res$mareyMap$map == "1")]
gen = res$mareyMap$gen[which(res$mareyMap$map == "1")]
# Re-map on a regular grid of evenly spaced markers
new = seq(50000, 30450000, by = 1000000)
smoothPar = res$smoothingParam[1]
newMap = predictGeneticMap(phys, gen, new, method = "loess", smoothing = smoothPar)

newMap
```


```{r newGeneticMap, message=FALSE, warning=FALSE, echo=FALSE, fig.height = 4, fig.width = 6, fig.align="center", fig.cap = "New positions on chromosome 1, with the confidence interval (+- 1.96*se)."}
newMap$phys = new
newMap = as.data.frame(newMap)

ggplot2::ggplot(data = newMap, aes(x = phys/10^6, y = newGen)) +
  geom_point(alpha = 0.4) +
  labs(x = "Genomic position (Mb)", y = "Genetic distance (cM)") +
  geom_errorbar(aes(ymin = newGen - 1.96*se, ymax = newGen + 1.96*se)) +
  theme(axis.line = element_line(),
            panel.grid.major = element_blank(),
            panel.grid.minor = element_blank(),
            panel.border = element_blank(),
            panel.background = element_blank(),
            plot.title = element_text(color="black", face="bold.italic",hjust = 0.5),
            plot.subtitle = element_text(color="black",hjust = 0.5),
            axis.title.x = element_text(color="black"),
            axis.title.y = element_text(color="black"),
            axis.text=element_text(colour="black"),
            legend.position = "right")
```





# References
