---
title: "How to estimate local recombination rates with the Marey map method"
output: rmarkdown::html_vignette
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{RecombinationMap}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The package is loaded with the following command.

```{r setup}
library(EasyMareyMap)
```

You can import the example dataset based on a linkage map in *Arabidopsis thaliana* [@serin_construction_2017].
The dataset need to be formatted according to the specifications of input files of the MareyMap package [@rezvoy_mareymap:_2007]. It is a tab-separated file with five columns named 'set, map, mkr, phys, gen, vld'. Genomic coordinates are in basepairs.


```{r importMap}
m = read.table("../data/Arabidopsis_thaliana.txt", header = T)
# The marey map is imported into a 'mareyMap' object
df = mareyMap(m)
# names of slots in the 'mareyMap' object
names(df)
# Column names of the input Marey map
names(df$mareyMap)
```

Here is the command to do outlier selection in a visual mode. You just need to draw a region around markers to remove. Left-click to place points of the polygon, then rigth-click to close the region and return to the script. All markers selected are automatically marked as removed (`vld =  FALSE`).

```
df = outlierSelection(df, "1")
```

<!-- TODO Add a photo of mkr selection -->


Alternatively, you can remove a whole chromosome of your dataset with the following command.


```{r removeChromosome}
df = removeChromosome(df, "5")
```


There is a method to `plot` the Marey map of a `mareyMap` object.

```{r plot, message=FALSE, warning=FALSE, echo=FALSE, fig.height = 4, fig.width = 6, fig.align="center", fig.cap = "Marey maps of five chromosomes in *Arabidopsis thaliana*. Invalid markers are shown in red."}
plot(df)
```

There is also a method to display a `summary` of a `mareyMap` object.

```{r summary}
summary(df)
```


All the procedure to estimate the recombination map is done with the `recombinationMap()` function which returns a `mareyMap` object with an estimated recombination map.

```{r estimateMap, echo=FALSE, message=FALSE}
res = recombinationMap(df, chromosome = "1", method = "loess")
res = recombinationMap(res, chromosome = "2", method = "loess")
res = recombinationMap(res, chromosome = "3", method = "spline")
```


```{r headMap, echo=TRUE, message=FALSE}
head(res$recMap)
```


```{r summaryResults, echo=TRUE, message=FALSE}
summary(res)
```

Finally, the recombination landscape can be plotted with the `recombinationPlot()` function.

```{r plotMap, message=FALSE, warning=FALSE, echo=FALSE, fig.height = 4, fig.width = 6, fig.align="center", fig.cap = "Recombination landscape of *Arabidopsis thaliana*. The gray area is the 95% confidence interval around the mean estimate."}
recombinationPlot(res)
```
